Kernel k-Groups via Hartigan's Method

### 2018-08-13 energy_clustering/pami18/paper-insert.tex
pami

\documentclass[10pt,journal,compsoc]{IEEEtran}

\usepackage{cite}
\usepackage{amsmath,amssymb,amsfonts,amsthm,amscd,bm,bbm}
\usepackage{mathtools}
%\mathtoolsset{showonlyrefs}
%\usepackage[inline]{enumitem}
\usepackage[pdftex]{graphicx}
\graphicspath{{./figs/}}
\usepackage{algorithmic}
\usepackage{algorithm}
%\usepackage[none]{hyphenat}
\usepackage{multirow}
\usepackage{colortbl}
\usepackage{array,booktabs}
\usepackage{xcolor}
\usepackage{makecell}
\usepackage{pifont}
\newcommand{\cmark}{\ding{51}}
\newcommand{\xmark}{\ding{55}}

\newtheorem{theorem}{Theorem}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{assumption}[theorem]{Assumption}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{conjecture}[theorem]{Conjecture}
\newtheorem{remark}[theorem]{Remark}
\newtheorem{example}{Example}

\DeclareMathOperator{\aff}{aff}
\DeclareMathOperator{\st}{s.t.}
\DeclareMathOperator{\affnot}{aff_0}
\DeclareMathOperator{\conv}{conv}
\DeclareMathOperator{\relint}{relint}
\DeclareMathOperator{\vol}{vol}
\DeclareMathOperator{\range}{range}
\DeclareMathOperator{\image}{im}
\DeclareMathOperator{\nullspace}{null}
\DeclareMathOperator{\area}{area}
\DeclareMathOperator{\vspan}{span}
\DeclareMathOperator{\id}{Id}
\DeclareMathOperator{\cond}{cond}
\DeclareMathOperator{\prox}{prox}
\DeclareMathOperator*{\argmax}{arg\,max}
\DeclareMathOperator*{\argmin}{arg\,min}
\DeclareMathOperator*{\minimize}{minimize}
\DeclareMathOperator{\diag}{diag}
\DeclareMathOperator{\Tr}{Tr}

\newcommand\Energy{\mathcal{E}}
\newcommand\EnergyH{\mathcal{E}^{H}}
\newcommand\EnergyL{\mathcal{E}^{L}}
\newcommand\EnergyOne{\mathcal{E}^{1D}}
\newcommand\E{\mathbb{E}}
\newcommand\kk{K}
\newcommand\kkk{h}
\newcommand\Hk{{\mathcal{H}}_{\kk}}
\newcommand\HH{\mathcal{H}}
\newcommand\C{{\mathcal{C}}}
\newcommand\tC{{\widetilde{\C}}}
\newcommand\OO{{\mathcal{O}}}
\newcommand\W{{\mathcal{W}}}
\newcommand\Zt{Y}
\newcommand{\Ind}[1]{\mathbbm{1}_{#1}}
\newcommand\e{e}
\newcommand\om{\omega}

\hyphenation{op-tical net-works semi-conduc-tor}

\begin{document}

\title{Kernel k-Groups via Hartigan's Method}

\author{Guilherme~Fran\c ca,~Maria Rizzo~and~Joshua T.~Vogelstein
\IEEEcompsocitemizethanks{\IEEEcompsocthanksitem G. Fran\c ca 
and J. Vogelstein are with the Center for Imaging Science, Johns Hopkins
University; J. Vogelstein is also with 
the Department of Biomedical Engineering and Institute for Computational
Medicine, Johns Hopkins University. \protect\\
% note need leading \protect in front of \\ to get a newline within \thanks as
% \\ is fragile and will error, could use \hfil\break instead.
E-mail: guifranca@jhu.edu; jovo@jhu.edu
\IEEEcompsocthanksitem M. Rizzo is with the Department of Mathematics
and Statistics, Bowling Green State University. \protect\\
E-mail: mrizzo@bgsu.edu}% <-this % stops an unwanted space
\thanks{Manuscript received March 15, 2018; revised March 15, 2018.}}

% The paper headers
\markboth{Energy Clustering}{}
%{Journal of \LaTeX\ Class Files,~Vol.~14, No.~8, August~2015}%

\IEEEtitleabstractindextext{%
\begin{abstract}
Energy statistics was proposed by Sz\' ekely in the 80's inspired by 
Newton's gravitational potential in classical mechanics, and it provides
a model-free 
hypothesis test for equality of distributions. In its original form, energy
statistics was formulated in 
Euclidean spaces. More recently, it 
was generalized to metric spaces of negative type. 
In this paper, we consider a formulation for the clustering problem
using a weighted version of 
energy statistics in spaces of negative type.
We show that this approach leads to a
quadratically constrained
quadratic program in the associated kernel space, establishing
connections with graph partitioning problems and
kernel methods in unsupervised machine learning.
To find local solutions of such an optimization problem, 
we propose an extension of Hartigan's method to kernel spaces.
Our method has the same computational cost
as kernel k-means algorithm, which is based on Lloyd's heuristic, but 
our numerical results show an improved performance,
especially in high dimensions.
\end{abstract}

% Note that keywords are not normally used for peerreview papers.
\begin{IEEEkeywords}
Clustering, Energy Statistics, Kernel Methods.
\end{IEEEkeywords}}

% make the title area
\maketitle

% To allow for easy dual compilation without having to reenter the
% abstract/keywords data, the \IEEEtitleabstractindextext text will
% not be used in maketitle, but will appear (i.e., to be "transported")
% here as \IEEEdisplaynontitleabstractindextext when the compsoc 
% or transmag modes are not selected <OR> if conference mode is selected 
% - because all conference papers position the abstract like regular
% papers do.
\IEEEdisplaynontitleabstractindextext
% \IEEEdisplaynontitleabstractindextext has no effect when using
% compsoc or transmag under a non-conference mode.

% For peer review papers, you can put extra information on the cover
% page as needed:
% \ifCLASSOPTIONpeerreview
% \begin{center} \bfseries EDICS Category: 3-BBND \end{center}
% \fi
%
% For peerreview papers, this IEEEtran command inserts a page break and
% creates the second title. It will be ignored for other modes.
\IEEEpeerreviewmaketitle

\IEEEraisesectionheading{\section{Introduction}\label{sec:introduction}}
% Computer Society journal (but not conference!) papers do something unusual
% with the very first section heading (almost always called "Introduction").
% They place it ABOVE the main text! IEEEtran.cls does not automatically do
% this for you, but you can achieve this effect with the provided
% \IEEEraisesectionheading{} command. Note the need to keep any \label that
% is to refer to the section immediately after \section in the above as
% \IEEEraisesectionheading puts \section within a raised box.

% The very first letter is a 2 line initial drop letter followed
% by the rest of the first word in caps (small caps for compsoc).
% 
% form to use if the first word consists of a single letter:
% \IEEEPARstart{A}{demo} file is ....
% 
% form to use if you need the single drop letter followed by
% normal text (unknown if ever used by the IEEE):
% \IEEEPARstart{A}{}demo file is ....
% 
% Some journals put the first two words in caps:
% \IEEEPARstart{T}{his demo} file is ....
% 
% Here we have the typical use of a "T" for an initial drop letter
% and "HIS" in caps to complete the first word.

\IEEEPARstart{E}{nergy Statistics} \cite{Szkely2013,Szkely2017}
is based on a 
notion of statistical potential energy between probability distributions,
in close analogy to Newton's gravitational potential in classical mechanics.
When probability distributions are different, the 
``statistical potential energy'' diverges as sample size increases, 
while tends 
to a nondegenerate limit distribution when probability
distributions are equal. 
Thus, it provides a model-free hypothesis test for equality of 
distributions which is achieved under minimum energy.

Energy statistics has been applied to several goodness-of-fit 
hypothesis tests, multi-sample tests of equality of distributions, 
analysis of variance \cite{RizzoVariance}, nonlinear dependence tests through
distance covariance and distance correlation~\cite{Szekely2007-mm}, which generalizes the Pearson
correlation coefficient, and hierarchical clustering 
by extending Ward's method of minimum variance
\cite{RizzoClustering}; 
see \cite{Szkely2013,Szkely2017} for an overview of energy
statistics and its applications.
Moreover, in Euclidean spaces, an application of 
energy statistics to clustering
was recently proposed \cite{Kgroups}, and the method was named 
\emph{k-groups}.

In its original formulation, energy statistics has a compact representation
in terms of expectations of pairwise Euclidean distances, providing
straightforward empirical estimates. 
More recently, the notion of distance covariance was further 
generalized from Euclidean 
spaces to metric spaces of negative type \cite{Lyons}. Furthermore, 
the link between energy distance based tests and kernel 
based tests has 
been recently established \cite{Sejdinovic2013} 
through an asymptotic equivalence between generalized energy distances and maximum
mean discrepancies (MMD), which are distances between embeddings of 
distributions in reproducing kernel Hilbert spaces (RKHS).
Even more recently, generalized energy distances and kernel methods have been demonstrated to be exactly equivalent, for all finite samples~\cite{Shen2018-st}.
This equivalence immediately relates energy statistics to
kernel methods often used in machine learning, and form the basis 
of our approach in this paper.

Clustering is an important unsupervised learning problem and 
has a long history in statistics and machine learning, making it
impossible to mention all important contributions in a short space. 
Perhaps, the most used method is k-means \cite{Lloyd,MacQueen,Forgy}, which
is based on Lloyd's heuristic \cite{Lloyd} of iteratively computing
the means of each cluster and then assigning points to
the cluster with closest center. The only statistical 
information about each cluster comes from its mean, making the method sensitive 
to outliers. Nevertheless, k-means works very well when data is 
linearly separable in Euclidean space. Gaussian mixture models (GMM) is 
another very common approach, providing more flexibility than k-means; 
however, it still makes strong assumptions about the distribution of 
the data.

To account for nonlinearities, kernel methods were introduced 
\cite{Smola,Girolami}. A Mercer kernel \cite{Mercer} is used to implicitly
map data points to a RKHS, then clustering can be performed in the associated
Hilbert space by using its inner product. However, the kernel choice remains 
the biggest challenge since there is no principled theory to construct a kernel
for a given dataset, and usually a kernel introduces hyperparameters that 
need to be carefully chosen. A well-known kernel based clustering method
is kernel k-means, which is precisely k-means 
formulated in the feature space \cite{Girolami}. 
Furthermore, kernel k-means algorithm
\cite{Dhillon2,Dhillon} is still based on Loyd's heuristic.
We refer the reader to \cite{Filippone} for a survey of clustering
methods.

Besides Lloyd's approach to clustering %, in Euclidean or kernel spaces,
there is an old heuristic
due to Hartigan \cite{Hartigan1975,Hartigan1979} that
goes as follows: for each data point, simply assign it to a cluster
in an optimal way such that a loss function is minimized.
While Lloyd's method only iterates if some cluster contains a point
that is closer to the mean of another cluster, Hartigan's method may iterate
even if that is not the case, and moreover, it takes into account the motion
of the means resulting from the reassignments. In this sense, Hartigan's
method may potentially escape local minima of Lloyd's method.
In the Euclidean case, this was shown to be the case \cite{Telgarsky}.
Moreover, the advantages of Hartigan's over Lloyd's method was verified
empirically \cite{Telgarsky,Slonin}. However, 
although it was observed to be as fast as Lloyd's method, no complexity
analysis was provided.

\subsection*{Contributions}
Although k-groups considers clustering from energy statistics
in the particular Euclidean case \cite{Kgroups},
the precise optimization problem behind this approach
remains obscure, as well as the connection with other methods
in machine learning.
The main theoretical contribution of this paper is to fill these gaps, 
which 
we do   more generality. For instance, our approach is not limited
to the Euclidean case but holds for general arbitrary spaces
of negative type. 
For generality, we also
consider a weighted version of energy statistics.
Our approach reveals connections
between energy statistics based clustering and existing methods
such as kernel k-means and graph partitioning problems.

Another contribution of this paper is to extend 
Hartigan's method to  kernel spaces. 
To the best of our knowledge, such an extension was not previously considered.
Since this approach was motivated by energy statistics and  
\cite{Kgroups} considered the Euclidean case, we call the 
proposed method \emph{kernel k-groups}.
We show that kernel k-groups 
has the same complexity as kernel k-means algorithm, however,
our numerical results provide compelling evidence that 
kernel k-groups 
is more accurate and robust, especially in high dimensions.

Using the standard kernel defined by energy
statistics, our experiments illustrate 
that kernel k-groups
is able to perform accurately on data coming from 
very different distributions, contrary to k-means and GMM, for instance.
More specifically, our method performs 
closely to k-means and GMM on normally distributed data, while
it is significantly better on data that 
is not normally distributed. 
Its superiority in high dimensions
is striking, being more accurate than k-means and GMM 
even in Gaussian settings. We also illustrate the advantages of
kernel k-groups on real data.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Review of Energy Statistics and RKHS}
\label{sec:background}

In this section, we introduce the main concepts from energy
statistics and its relation to 
RKHS which form the basis of our work.
For more details we refer 
to \cite{Szkely2013} and \cite{Lyons,Sejdinovic2013}.

Consider random variables in $\mathbb{R}^D$ 
such that $X,X' \stackrel{iid}{\sim} P$ and 
$Y,Y' \stackrel{iid}{\sim} Q$, where $P$ and $Q$ are cumulative
distribution functions with finite first moments. 
The quantity 
\begin{equation}
\label{eq:energy}
\Energy(P, Q) \equiv 2 \E \| X - Y\| - \E \| X - X' \| - \E \| Y - Y' \|,
\end{equation}
called \emph{energy distance} \cite{Szkely2013}, 
is rotationally invariant and nonnegative, $\Energy(P,Q) \ge 0$, where
equality
to zero holds if and only if $P = Q$.
Above, $\| \cdot \|$ denotes the
Euclidean norm in $\mathbb{R}^D$. 
Energy distance
provides a characterization of equality of distributions, and
$\Energy^{1/2}$ is
a metric on the space of distributions.

The energy distance can be generalized as, for instance,
\begin{equation}
\label{eq:energy2}
\Energy_\alpha(P, Q) \equiv 
2 \E \| X - Y\|^{\alpha} - \E \| X - X' \|^{\alpha} - 
\E \| Y - Y' \|^{\alpha}
\end{equation}
where $0<\alpha\le 2$. This quantity is also nonnegative,
$\Energy_\alpha(P,Q) \ge 0$. Furthermore, for $0<\alpha<2$ we have that
$\Energy_\alpha(P,Q) = 0$ if and only if $P=Q$, while for $\alpha=2$ 
we have $\Energy_2(P,Q) = 2\| \E(X) - \E(Y) \|^2$ which shows that
equality to zero only requires
equality of the means, and thus $\Energy_2(P,Q)=0$ does 
not imply equality of distributions.

The energy distance can be even further generalized.
Let $X, Y \in \mathcal{X}$  where $\mathcal{X}$ is an arbitrary space endowed
with a \emph{semimetric of negative type}
$\rho: \mathcal{X}\times\mathcal{X} \to \mathbb{R}$, which is required
to satisfy
\begin{equation}
\label{eq:negative_type}
\sum_{i,j=1}^n c_i c_j \rho(X_i, X_j) \le 0,
\end{equation}
where $X_i \in \mathcal{X}$ and $c_i \in \mathbb{R}$ such that
$\sum_{i=1}^n c_i = 0$. Then, $\mathcal{X}$ is called a \emph{space of
negative type}.
We can thus replace $\mathbb{R}^D$ by $\mathcal{X}$ and 
$\| X - Y \|$ by $\rho(X , Y)$ in the definition \eqref{eq:energy}, obtaining
the \emph{generalized energy distance}
\begin{equation}
\label{eq:energy3}
\Energy(P, Q) \equiv 2 \E \rho(X,Y) - \E \rho(X, X') - \E \rho(Y,Y').
\end{equation}
For spaces of negative type, there exists a Hilbert space $\mathcal{H}$ and
a map $\varphi: \mathcal{X} \to
\mathcal{H}$ such that
$\rho(X, Y) = \| \varphi(X) - \varphi(Y) \|_{\mathcal{H}}^2$. This
allows us to compute quantities related to probability distributions over
$\mathcal{X}$ in the associated Hilbert space $\mathcal{H}$.
Even though the semimetric 
$\rho$ may not satisfy the triangle inequality, 
$\rho^{1/2}$ does since it can be shown to be a proper metric. 
Our energy clustering formulation, proposed in the next section,
will be based on the generalized
energy distance \eqref{eq:energy3}.

There is an equivalence between energy distance, 
commonly used in statistics,
and distances between embeddings of distributions in 
RKHS, commonly used in machine learning. 
This equivalence was established
in \cite{Sejdinovic2013}. Let us first recall the definition of
RKHS. Let $\HH$ be a Hilbert space of real-valued functions
over $\mathcal{X}$. A function 
$\kk : \mathcal{X} \times \mathcal{X} \to 
\mathbb{R}$ is a reproducing kernel of $\HH$ if it satisfies
the following two conditions:
\begin{enumerate}
\item $\kkk_x \equiv \kk(\cdot, x) \in \HH$ 
for all $x \in \mathcal{X}$;
\item $\langle \kkk_x, f \rangle_{\HH} = f(x)$ for
all $x\in\mathcal{X}$ and $f\in \HH$.
\end{enumerate}
In other words, for any $x \in \mathcal{X}$ and any function $f \in \HH$,
there is a unique 
$\kkk_x \in \HH$ that reproduces $f(x)$ through the inner product
of $\HH$.
If such a \emph{kernel} 
function $\kk$ exists, then $\HH$ is called a RKHS. The above two 
properties immediately imply that $\kk$ is symmetric and positive
definite. 
Defining the Gram matrix $G$ with
elements $G_{ij} = \kk(x_i,x_j)$, this is equivalent to $G=G^\top$ being
positive semidefinite, i.e., $v^\top G \, v \ge 0$ for any vector
$v \in \mathbb{R}^n$.

The Moore-Aronszajn theorem 
\cite{Aronszajn}
establishes the converse of the above paragraph.
For every symmetric
and positive definite function $\kk: \mathcal{X}\times \mathcal{X} \to
\mathbb{R}$, there is an associated RKHS, $\Hk$,
with reproducing
kernel $\kk$. The map $\varphi: x \mapsto \kkk_x \in \Hk$ is called
the canonical \emph{feature map}. Given a kernel $\kk$,
this theorem enables us to define an embedding of a probability measure
$P$ into the RKHS as follows: $P \mapsto \kkk_P \in
\Hk$ such that 
$\int f(x) d P(x) = \langle f, \kkk_P \rangle$ for all $f \in \Hk$,
or alternatively, $\kkk_P \equiv \int \kk( \, \cdot \,, x)  d P(x)$. 
We can now  introduce the 
notion of distance between two probability measures using the inner product
of $\Hk$, which is called the maximum mean discrepancy (MMD) and
is given by
\begin{equation}
\label{eq:mmd}
\gamma_\kk(P,Q) \equiv \| \kkk_P - \kkk_Q \|_{\Hk}.
\end{equation}
This can also be written as \cite{Gretton2012}
\begin{equation}\label{eq:mmd2}
\gamma_\kk^2(P,Q) = \E \kk(X,X') + \E \kk(Y,Y') - 2 \E \kk(X, Y)
\end{equation}
where $X,X' \stackrel{iid}{\sim} P$ and $Y,Y'\stackrel{iid}{\sim} Q$.
From the equality between \eqref{eq:mmd} and \eqref{eq:mmd2} we also
have $\langle \kkk_P, \kkk_Q \rangle_{\Hk} = \E \, \kk(X, Y)$.

The following important result shows that semimetrics of negative
type and symmetric positive definite kernels are closely related
\cite{Berg1984}. Let $\rho: \mathcal{X} \times \mathcal{X} \to \mathbb{R}$
and $x_0 \in \mathcal{X}$ an arbitrary but fixed point.
Define
\begin{equation}
\label{eq:kernel_semimetric}
\kk(x,y) \equiv 
\tfrac{1}{2} \left[  \rho(x,x_0) + \rho(y,x_0) - \rho(x,y)\right].
\end{equation}
Then, it can be shown that 
$\kk$ is positive definite if and only if $\rho$ is a semimetric
of negative type.
We have a family of kernels, one for each choice of $x_0$. Conversely,
if $\rho$ is a semimetric of negative type and $\kk$ is a kernel in this
family, then 
\begin{equation}
\label{eq:gen_kernel}
\begin{split}
\rho(x,y) &= \kk(x,x) + \kk(y,y) -2\kk(x,y) \\
&=  \| \kkk_x - \kkk_y \|^2_{\Hk}
\end{split}
\end{equation}
and the canonical feature map 
$\varphi: x \mapsto \kkk_x$ is injective \cite{Sejdinovic2013}.
When these conditions are satisfied, we say that the kernel $\kk$ 
generates the semimetric $\rho$. 
If two different kernels generate the same $\rho$, they are
said to be equivalent kernels.

Now we can state the equivalence between the generalized 
energy distance \eqref{eq:energy3} and
inner products on RKHS, which is one of the main results of
\cite{Sejdinovic2013}. If $\rho$ is a semimetric
of negative type and $\kk$ a kernel that generates $\rho$, then
replacing \eqref{eq:gen_kernel} into
\eqref{eq:energy3}, and using \eqref{eq:mmd2}, yields
\begin{equation} \label{eq:Erho}
\begin{split}
\Energy(P, Q) &= 
2 \left[ \E \, \kk(X, X') + \E \, \kk(Y, Y') - 2\E \, \kk(X, Y)\right]  \\
&= 2 \gamma_\kk^2(P,Q) .
\end{split}
\end{equation}
Due to \eqref{eq:mmd}, we can compute the energy 
distance $\mathcal{E}(P, Q)$ between two probability distributions
using the inner 
product of $\Hk$.

Finally, let us recall the main formulas from generalized energy statistics
for the test statistic of equality of distributions \cite{Szkely2013}. 
Assume that we have data $\mathbb{X} = \{ x_1,\dotsc, x_n \}$, where
$x_i \in \mathcal{X}$, and $\mathcal{X}$ is a space of negative type.
Consider a disjoint partition $\mathbb{X} = \bigcup_{j=1}^k \C_j$, with
$\C_i \cap \C_j = \emptyset$.
Each expectation in the generalized energy distance
\eqref{eq:energy3}
can be computed 
through the function
\begin{equation}
\label{eq:g_def}
g (\C_i, \C_j) \equiv 
\dfrac{1}{n_i n_j}
\sum_{x \in \C_i} 
\sum_{y \in \C_j} \rho(x, y) ,
\end{equation}
where $n_i = |\C_i|$ is the number of elements in partition
$\C_i$. 
The \emph{within energy dispersion} is defined by
\begin{equation}
\label{eq:within}
W \equiv
\sum_{j=1}^{k} \dfrac{n_j}{2} g(\C_j, \C_j),
\end{equation}
and the \emph{between-sample energy statistic} is defined by
\begin{equation}
\label{eq:between}
S \equiv \!\!\!\!
\sum_{1 \le  i < j \le k } \dfrac{n_i n_{j}}{2 n} \left[
2 g(\C_i, \C_j) - 
g(\C_i, \C_i) - 
g(\C_j, \C_j)
\right],
\end{equation}
where $n = \sum_{j=1}^k n_j$.
Given a set of distributions
$\{ P_j\}_{j=1}^k$, where $x \in \C_j$ if and only if $x \sim P_j$, 
the quantity $S$ provides
a test statistic for equality of distributions
\cite{Szkely2013}.
When the sample size is large enough, $n\to \infty$,
under the null hypothesis $H_0: P_1=P_2=\dotsm=P_k$, we have that
$S\to 0$, 
and under
the alternative hypothesis $H_1: P_i \ne P_j$ for at least two $i\ne j$, 
we have that $S \to \infty$.
%Note that this test does not make any assumptions
%about the form of distribution $P_j$, thus it is said to be 
%distribution-free.

%One can make a physical analogy by thinking 
%that points $ x \in \C_j$ form a massive body 
%whose total mass is characterized by the distribution function $P_j$.
%The quantity $S$ is thus a potential
%energy of the from $S(P_1,\dotsc,P_k)$ which measures how different
%the distribution of these masses are,  and achieves the ground state
%$S=0$ when all bodies have the same mass distribution. On the other hand,
%the potential energy
%$S$ increases as bodies have different mass distributions.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{The Clustering Problem Formulation}
\label{sec:clustering_theory}

This section contains our main theoretical results. First, we
generalize the previous formulas from energy statistics by introducing
weights associated to data points. 
Second, we formulate an optimization problem for clustering 
in the associated RKHS, making connection with kernel methods in
machine learning.

Let $w(x)$ be a weight function associated to point $x \in \mathcal{X}$
and define
\begin{equation}
\label{eq:g_def2}
g(\C_i, \C_j) \equiv \dfrac{1}{s_i s_j} \sum_{x\in \C_i}\sum_{y\in\C_j}
w(x)w(y) \rho(x,y),
\end{equation}
where
\begin{equation}
s_i \equiv \sum_{x\in\C_i} w(x), \qquad
s \equiv \sum_{j=1}^k s_j.
\end{equation}
The  weighted version of the within energy dispersion 
and between-sample
energy statistic are thus given by
\begin{align}
W &\equiv
\sum_{j=1}^{k} \dfrac{s_j}{2} g(\C_j, \C_j), \label{eq:within2} \\
S &\equiv \!\!\! \sum_{1 \le  i < j \le k } \!\!\! 
\dfrac{s_i s_{j}}{2 s} \left[
2 g(\C_i, \C_j) - 
g(\C_i, \C_i) - 
g(\C_j, \C_j)
\right]. \label{eq:between2}
\end{align}
Note that if $w(x) = 1$ for every $x$ we recover the previous formulas.

Due to the test statistic for equality of distributions,
the obvious criterion for clustering data is to 
maximize $S$ in \eqref{eq:between2},
which makes  each cluster as different as possible from the other ones.
In other words, given a set of points coming from different probability
distributions, the test statistic $S$ should attain a maximum when 
each point is correctly classified as belonging to the cluster associated 
to its probability distribution. The following 
result
shows that maximizing $S$ is, however, equivalent to minimizing
$W$ in \eqref{eq:within2}.% which has a more convenient form.

\begin{lemma}
\label{th:minimize}
Let $\mathbb{X} = \{x_1,\dotsc,x_n\}$ where each data point
$x_i$ lives in a space $\mathcal{X}$ endowed with a semimetric $\rho:
\mathcal{X}\times\mathcal{X} \to \mathbb{R}$ of
negative type. For a fixed integer $k$,
the partition
$\mathbb{X} = \bigcup_{j=1}^k \C^\star_j$, where 
$\C^\star_i \cap C^\star_j = \emptyset$ for
all $i\ne j$, maximizes the between-sample statistic $S$, defined
in equation \eqref{eq:between2}, if and only if
\begin{equation}
\label{eq:minimize}
\{ \C_1^\star,\dotsc,\C_k^\star \} = \argmin_{\C_1,\dotsc,C_k  } W(
\C_1, \dotsc, \C_k) ,
\end{equation}
where the within energy dispersion $W$ is defined by \eqref{eq:within2}.
\end{lemma}
\begin{proof}
From \eqref{eq:within2} and \eqref{eq:between2}
we have that
\begin{align}
%\begin{equation}
%\begin{split}
& S + W  
\nonumber \\
&= \dfrac{1}{2s} \sum_{\substack{i,j=1 \\ i\ne j}}^k s_i s_j g(\C_i, \C_j)
+ \dfrac{1}{2s} \sum_{i=1}^{k} 
\bigg[ s - 
\sum_{\substack{j= 1 \\ j\ne i}}^k s_j \bigg] 
s_i g(\C_i, \C_i) 
\nonumber \\
& = \dfrac{1}{2s} \sum_{i,j=1}^k s_i s_j g(\C_i, \C_j) 
\nonumber \\
& = \dfrac{1}{2s} \sum_{x \in \mathbb{X}} \sum_{y \in \mathbb{X}} 
w(x) w(y)\rho(x,y)  
\nonumber \\
&= \dfrac{s}{2} g(\mathbb{X}, \mathbb{X}).
%\end{split}
%\end{equation}
\end{align}
Since $g(\mathbb{X}, \mathbb{X})$ is independent 
of the choice of partition, $\max_{\{ \C_i \}} S = - \max_{\{\C_i\}} W = 
\min_{\{ \C_i \}}W$, as claimed.
\end{proof}

For a given $k$, the clustering problem amounts to
finding the best partitioning of the data by minimizing $W$.
In the current form of problem \eqref{eq:minimize},
the relationship
with other clustering methods or kernel spaces is totally obscure.
In the following, we demonstrate what is the explicit 
optimization problem behind 
\eqref{eq:minimize} in the corresponding RKHS, 
which establishes the connection with kernel methods.

Based on the relationship between kernels and semimetrics of negative
type, 
assume that the kernel $\kk: \mathcal{X} \times \mathcal{X} \to \mathbb{R}$ 
generates $\rho$.  Define  the Gram matrix
\begin{equation}
\label{eq:kernel_matrix}
G \equiv \begin{pmatrix}
\kk(x_1,x_1)     & \dotsm     & \kk(x_1,x_n) \\
\vdots           & \ddots     & \vdots \\
\kk(x_n,x_1)     & \dotsm     & \kk(x_n,x_n) 
\end{pmatrix} .
\end{equation}
Let $Z \in \{ 0,1 \}^{n\times k}$ be the label matrix, 
with only one nonvanishing entry per row, 
indicating to which cluster (column)
each point (row) belongs to. This matrix satisfies
$Z^\top Z = D$, where the diagonal matrix 
$D = \diag( n_1,\dotsc, n_k )$  contains
the number of points in each cluster. We also introduce the rescaled
matrix  $Y$ below. In component form they are 
\begin{equation}
\label{eq:label_matrix}
Z_{ij} \equiv \begin{cases}
1 & \mbox{if $x_i \in \C_j$ } \\
0 & \mbox{otherwise}
\end{cases}, \qquad
\Zt_{ij} \equiv \begin{cases}
\tfrac{1}{\sqrt{s_j}} & \mbox{if $x_i \in \C_j$ } \\
0 & \mbox{otherwise}
\end{cases} .
\end{equation}
Throughout the paper, we use the notation $M_{i\bullet}$ to denote
the $i$th row of a matrix $M$, and $M_{\bullet j}$ denotes its $j$th column.
We also define the following:
\begin{equation}
\label{eq:weighted_matrices}
\mathcal{W} \equiv \diag(w_1,\dotsc,w_n), \quad
H \equiv \mathcal{W}^{1/2} Y, \quad
\om \equiv \mathcal{W} \e,
\end{equation}
where $w_i = w(x_i)$ is the weight associated to point $x_i$, 
and $\e =(1,\dotsc,1)^\top \in \mathbb{R}^n$ 
is the all-ones vector.

Our next result shows that the optimization problem \eqref{eq:minimize}
is NP-hard since it is a quadratically constrained quadratic program 
(QCQP) in the associated RKHS.

\begin{theorem}
\label{th:qcqp3}
The optimization 
problem \eqref{eq:minimize} is equivalent to
\begin{equation}
\label{eq:qcqp3}
\begin{split}
&\max_H  \Tr \left[ H^\top \left(\mathcal{W}^{1/2} \, G \, 
\mathcal{W}^{1/2}\right) H  \right] \\
&\mbox{such that $H \ge 0$, $H^\top H = I$, $H H^\top \om = \om$,}
\end{split}
\end{equation}
where $G$ is the Gram matrix \eqref{eq:kernel_matrix} 
and the other quantities are defined in \eqref{eq:weighted_matrices}.
\end{theorem}
\begin{proof}
From 
\eqref{eq:gen_kernel},
\eqref{eq:g_def2}, and
\eqref{eq:within2}
we have
\begin{equation}
\label{eq:W2}
\begin{split}
\!\! \!\!     W
&= \sum_{j=1}^k \dfrac{1}{2 s_j} 
\!\sum_{x,y \in \C_j} w(x)w(y)\rho(x,y) \\
&= \! \sum_{j=1}^k   \sum_{x \in \C_j} \!\! \bigg[
w(x)\kk(x,x) \!  - \! \dfrac{1}{s_j} \! \sum_{y \in \C_j} \!w(x)w(y) \kk(x,y) \bigg].
\end{split}
\end{equation}
Note that the first term is global so it does not contribute to the 
optimization problem.
Therefore, problem \eqref{eq:minimize} becomes
\begin{equation}
\label{eq:max_prob}
\max_{ \C_1,\dotsc,\C_k } 
\sum_{j=1}^k \dfrac{1}{s_j} \sum_{x,y\in C_j} w(x) w(y) \kk(x,y) .
\end{equation}
Using the definitions 
\eqref{eq:label_matrix} and 
\eqref{eq:weighted_matrices}, the previous 
objective function can be written as
\begin{equation}
\begin{split}
& \hspace{-2em} \sum_{j=1}^k \dfrac{1}{s_j} 
\sum_{p=1}^n \sum_{q=1}^n 
w_p w_q Z_{pj} Z_{qj} G_{pq}  \\
&= 
\sum_{j=1}^k 
\sum_{p=1}^n \sum_{q=1}^n 
\dfrac{Z^\top_{jp}\sqrt{w_p}}{\sqrt{s_j}} w_p^{1/2} G_{pq} w_q^{1/2} 
\dfrac{\sqrt{w_q} Z_{qj}}{\sqrt{s_j}} \\
&= 
\sum_{j=1}^k \left(H^\top \mathcal{W}^{1/2} G \mathcal{W}^{1/2} H\right)_{jj}
\\
&= \Tr\left[ H^\top \mathcal{W}^{1/2} \, G \, \mathcal{W}^{1/2} H  \right].
\end{split}
\end{equation}

>>>>

%%%%%%%% insert above Conclusion???

An implementation of unweighted $k$-groups is available in the
energy package for R \cite{energy}.

%%%%%%%%%  here is the citation for the package, update submitted to CRAN on Aug. 11

  Maria Rizzo and Gabor Szekely (2018). energy: E-Statistics: Multivariate Inference via the Energy of
  Data. R package version 1.7-5. https://CRAN.R-project.org/package=energy

or

@Manual{energy,
    title = {energy: E-Statistics: Multivariate Inference via the Energy of Data},
    author = {Maria Rizzo and Gabor Szekely},
    year = {2018},
    note = {R package version 1.7-5},
    url = {https://CRAN.R-project.org/package=energy},
  }

%%%%%%%%%%% bio sketch for Maria inserted below %%%%%%%%%%%%
